Individual and Differential Harm in Redistricting

### 2025-08-25 paper/harm.qmd
typos

We stress that the results here are tied the specific counterfactual plan (b).
Indeed, this is a feature of the harm framework: different research questions may require different counterfactuals.
We illustrate this aspect of the framework throughout the applications below.

@@

To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a (random) partition of the set of voters.
In general we do require the group identities to be fixed with respect to elections.[^1]
This is because some group identities, like partisan preference, may themselves change election-to-election.
Groups may need to be estimated with methods such as ecological inference [@king1997solution], as discussed in @sec-estimating.

@@

Once we have defined a set of groups, we define the *differential harm* for two groups $G_k$ and $G_l$ as
$$
  \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l].
$$
So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.

Zero differential harm between groups is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and negative rates across groups [@zafar2017fairness; @chouldechova2017fair].
There, the implicit counterfactual is a perfect predictor and the actual process harms individuals whenever it makes an incorrect prediction.
Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.

@@

For differential harm to measure gerrymandering, a counterfactual $\tilde Q$ drawn without partisan intent is needed.
To do so, we generate 10,000 districting plans without any partisan data, using the SMC algorithm of [@mccartan2020; implemented in @redist] setting constraints to favor compact plans that respect county lines.
@fig-nj-maps(d) shows one such plan.
For each of these sampled plans and the two proposals, we estimate the average harm and differential partisan harm versus the simulated baseline,[^9] and calculate the expected number of Democratic seats and several common partisan metrics, including partisan dislocation.
These measures, and their distribution in the basline, are shown in @fig-nj-meas.

@@

One limitation of the harm definition itself is that considers only individuals who vote; abstainers experience zero harm.
A richer framework and the use of auxiliary survey data might allow consideration of the informal preferences of the whole population, even when those preferences are not expressed at the ballot box.

@@

Motivated primarily by computational convenience and simplicity, we recommend for most cases the stochastic uniform partisan swing model of @gelman1994unified, modeling the Democratic two-party vote share in district $d$ and election $v$ as
$$
    s_{dv} = \bar{s}_d + \delta_v + \varepsilon_{dv},
$$
where $\bar{s}_d$ is the baseline Democratic vote share in the districts, $\delta_v\sim\Norm(0, \sigma^2_\delta)$ is a election-specific uniform shift, and $\varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon)$ is a district-election residual error term.
This can be thought of as a hiererchical extension of the classic uniform swing model.

@@

A quite different approach to measuring partisan gerrymandering from symmetry-based methods is the partisan dislocation [@defoeubarodd21].
This measure is based on a $k$-nearest-neighbor comparison, where $k$ is the size of a district, to identify when a voter's district is politically different from their assigned district.
This can be repeated for  voter in a state or region, demonstrating at a local level when people been placed in a district which is atypical for who their nearest neighbors are.
When aggregated to the precinct-level or beyond, this can produce local or statewide measures of partisan gerrymandering, similar to our proposed differential partisan harm measure.

@@

Despite these differences, there are also empirical similarities in many cases.
The efficiency gap is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates differential partisan harm as well.
Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in @sec-nj.

@@

```{r fig-alabama-harm-sample, fig.width=6.5}
#| fig.cap: |
#|      On the left, the expected fraction of voters in each precinct which are
#|      harmed by the 2020 enacted congressional districts versus the sampled
#|      counterfactual set.
#|      On the right, the expected number of harmed voters in each race/partisan
#|      group, relative to the total number of such voters.
include_graphics("figures/al_harm_sample.pdf")
```
@fig-alabama-harm-sample summarizes the results, which are mostly similar to the results in in @sec-alabama.
Across the state, the estimated average harm is 6.9%, which translates to around 144,000 voters.
Of these harmed voters, around 56,000 are Black, while 88,000 are White.
Similarly, 73,000 are Democratic, while only 71,000 are Republican.
Normalizing by group sizes, the expected harm for a Black Democrat is 9.6%, for a Black Republican is 16.8%, for a White Democrat is 9.7%, and for a White Republican is 5.2%.
So the challenged plan differentially harms Democratic voters, but does not differentially harm Black Democrats versus White Democrats.
However, it it especially harms Black Republicans, who have an estimated differential harm of 7.2% versus Black Democrats and of 11.6% versus White Republicans.

>>>>

We stress that the results here are tied to the specific counterfactual plan (b).
Indeed, this is a feature of the harm framework: different research questions may require different counterfactuals.
We illustrate this aspect of the framework throughout the applications below.

@@

To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a (random) partition of the set of voters.
In general we will not require the group identities to be fixed with respect to elections.[^1]
This is because some identities, like partisan preference, may themselves change election-to-election.

@@

Once we have defined a set of groups, we define the *differential harm* for two groups $G_k$ and $G_l$ as
$$
  \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l].
$$
So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.
The estimation of harm by group may require methods like ecological inference [@king1997solution], as discussed in @sec-estimating.

Zero differential harm between groups is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and negative rates across groups [@zafar2017fairness; @chouldechova2017fair].
There, the implicit counterfactual is a perfect predictor; the actual process harms individuals whenever it makes an incorrect prediction.
Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.

@@

For differential harm to measure gerrymandering, a counterfactual $\tilde Q$ drawn without partisan intent is needed.
To do so, we generate 10,000 districting plans without any partisan data, using the SMC algorithm of [@mccartan2020; implemented in @redist] setting constraints to favor compact plans that respect county lines.
@fig-nj-maps(d) shows one such plan.
For each of these sampled plans and the two proposals, we estimate the average harm and differential partisan harm versus the simulated baseline,[^9] and calculate the expected number of Democratic seats and several common partisan metrics, including partisan dislocation.
These measures, and their distribution in the baseline, are shown in @fig-nj-meas.

@@

One limitation of the harm definition is that it considers only voters; abstainers experience zero harm.
A richer framework and the use of auxiliary survey data might allow consideration of the informal preferences of the whole population, even when those preferences are not expressed at the ballot box.

@@

Motivated primarily by computational convenience and simplicity, we recommend for most cases the stochastic uniform partisan swing model of @gelman1994unified, modeling the Democratic two-party vote share in district $d$ and election $v$ as
$$
    s_{dv} = \bar{s}_d + \delta_v + \varepsilon_{dv},
$$
where $\bar{s}_d$ is the baseline Democratic vote share in the districts, $\delta_v\sim\Norm(0, \sigma^2_\delta)$ is a election-specific uniform shift, and $\varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon)$ is a district-election residual error term.
This can be thought of as a hierarchical extension of the classic uniform swing model.

@@

A quite different approach to measuring partisan gerrymandering from symmetry-based methods is the partisan dislocation [@defoeubarodd21].
This measure is based on a $k$-nearest-neighbor comparison, where $k$ is the size of a district, to identify when a voter's district is politically different from their assigned district.
This can be repeated for each voter in a state or region, demonstrating at a local level when people have been placed in a district which is atypical for who their nearest neighbors are.
When aggregated to the precinct-level or beyond, this can produce local or statewide measures of partisan gerrymandering, similar to our proposed differential partisan harm measure.

@@

Despite these differences, there are also empirical similarities in many cases.
The efficiency gap is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates of differential partisan harm as well.
Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in @sec-nj.

@@

```{r fig-alabama-harm-sample, fig.width=6.5}
#| fig.cap: |
#|      On the left, the expected fraction of voters in each precinct which are
#|      harmed by the 2020 enacted congressional districts versus the sampled
#|      counterfactual set.
#|      On the right, the expected number of harmed voters in each race/partisan
#|      group, relative to the total number of such voters.
include_graphics("figures/al_harm_sample.pdf")
```
@fig-alabama-harm-sample summarizes the results, which are mostly similar to the results in @sec-alabama.
Across the state, the estimated average harm is 6.9%, which translates to around 144,000 voters.
Of these harmed voters, around 56,000 are Black, while 88,000 are White.
Similarly, 73,000 are Democratic, while only 71,000 are Republican.
Normalizing by group sizes, the expected harm for a Black Democrat is 9.6%, for a Black Republican is 16.8%, for a White Democrat is 9.7%, and for a White Republican is 5.2%.
So the challenged plan differentially harms Democratic voters, but does not differentially harm Black Democrats versus White Democrats.
However, it it especially harms Black Republicans, who have an estimated differential harm of 7.2% versus Black Democrats and of 11.6% versus White Republicans.
